---
title: "Habitat Suitability for Aquaculture"
author: "Carmen Hoyt"
date: last-modified
execute: 
  warning: false
  messages: false
format:
  html:
    toc: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---
# About

Aquaculture has the potential to match the production of wild fisheries using just a fraction of ocean space: 0.015% of the global ocean area, to be precise[^1]. To make efficient use of this space, it is essential to match species of farming potential with their optimal habitat for growth. Many variables influence habitat suitability, but this analysis focused on two main factors: sea surface temperature (SST) and depth. There are 5 main Exclusive Economic Zones (EEZs) off the West Coast of the US: Washington, Oregon, Northern California, Central California, and Southern California. By creating a `suitability()` function, we can determine how much area in each EEZ is suitable for aquaculture for any species of interest[^2].

[^1]: Read more about the findings of [Gentry el al.](https://www.nature.com/articles/s41559-017-0257-9).

[^2]: Temperature and depth preferences can be obtained for many marine species from [SeaLifeBase](https://www.sealifebase.ca/search.php).

# Data

All the data used in this analysis were downloaded from a pre-prepared [Google Drive](https://drive.google.com/file/d/1u-iwnPDbe6ZK7wSFVMI-PpCKaRQ3RVmg/view). The data in the drive was previously accessed from the following sources:

**Sea Surface Temperature (SST) Data:**

SST data was originally obtained from [NOAA's Daily Global 5km Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php) from the years 2008 to 2012 as a GeoTiff. 

**Bathymetry Data:**

Bathymetry data was originally downloaded from the [General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area) as a GeoTiff.

**EEZ Data:**

EEZ boundaries were downloaded as a shapefile from [Marineregions.org](https://www.marineregions.org/eez.php).

# Analysis

## Import packages
```{r}
#| code-summary: Expand code
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(dplyr)
library(kableExtra)
library(terra)
library(patchwork)
library(tmaptools)
```

## Read in data
```{r}
#| code-summary: Expand code
# Read in EEZ data
west_coast_eez <- read_sf(here::here("data", "wc_regions_clean.shp"))

# Read in and rasterize bathymetry data
bathymetry <- stars::read_stars(here::here("data", "depth.tif"))
bathymetry <- rast(bathymetry)

# Read in and rasterize SST (sea surface temp) data
sst_rasters <- list.files(here("data"), pattern = glob2rx("average_annual*"), full.names = TRUE)

for (i in seq_along(sst_rasters)) {
  raster <- stars::read_stars(sst_rasters[i])
  raster <- rast(raster)
  raster_name <- paste0("sst", str_sub(sst_rasters[i], -9 ,-5))
  if (st_crs(west_coast_eez) == st_crs(raster)) {
  print("The CRS match! Good to go.")
  } else {
  raster <- project(raster, crs(west_coast_eez))
  }
  try(if(st_crs(west_coast_eez) != st_crs(raster)) stop("CRS doesn't match."))
  assign(raster_name, raster)
}

# Stack SST rasters
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)

# Find mean SST
avg_sst <- mean(sst, na.rm = TRUE)

# Convert from Kelvin to Celsius
avg_sst <- avg_sst - 273.15

# Clip bathymetry to same extent as SST
depth <- bathymetry[avg_sst, drop = FALSE]

# Resample depth to match SST resolution
depth_resample <- terra::resample(depth, y = avg_sst, method = "near")

# Stack SST and depth rasters
stack <- c(sst, depth_resample)
```

## Suitability Function

The function `suitability()` takes the following arguments (in order):

- `species_name`: the name of the species (as a string)

- `min_temp`: minimum sea surface temperature (degrees Celsius)

- `max_temp`: maximum sea surface temperature (degrees Celsius)

- `min_depth`: the absolute value of the shallowest preferred depth (meters)

- `max_depth`: the absolute value of the deepest preferred depth (meters)

The final product is a map of West Coast EEZ regions colored by area (km^2) associated with suitable growing conditions for the species of interest. 

Example: `suitability("Abalone", 8, 18, 0, 24)` where:

- `species_name`: "Abalone"

- `min_temp`: 8 (degrees C)

- `max_temp`: 18 (degrees C)

- `min_depth`: 0 (meters)

- `max_depth`: 24 (meters)

```{r}
#| code-summary: Expand code
# Create suitability function
suitability <- function(species_name, min_temp, max_temp, min_depth, max_depth) {
  
  # Save species name
  species_name = species_name
  
  # Reclassify temp raster for suitable temperatures
  rcl_temp <- matrix(c(-Inf, min_temp, 0, # min, max, group
               min_temp, max_temp, 1,
               max_temp, Inf, 0),
             ncol = 3, byrow = TRUE)

  sst_suitable <- terra::classify(avg_sst, rcl = rcl_temp)

  # Reclassify depth raster for suitable depths
  rcl_depth <- matrix(c(-Inf, -max_depth, 0, # min, max, group
               -max_depth, -min_depth, 1,
               -min_depth, Inf, 0),
             ncol = 3, byrow = TRUE)

  depth_suitable <- terra::classify(depth_resample, rcl = rcl_depth)

  # Stack the two reclassified temp and depth rasters and multiply to find where suitability overlaps 
  fun <- function(x,y){return(x*y)}
  stack <- c(sst_suitable, depth_suitable)
  suitable_raster <- lapp(stack, fun)

  # Replace 0 values with NA
  suitable_raster[suitable_raster == 0] <- NA
  
  # Find suitable area masked to US EEZ extent
  suitable_raster <- mask(suitable_raster, west_coast_eez)
  
  # Rasterize the EEZ to calculate area
  eez_raster <- rasterize(west_coast_eez, stack, "rgn")
  
  # Find the size of each cell in suitable_raster
  cell_size <- cellSize(suitable_raster, unit = "km")
  
  # Calculate total suitable area for each EEZ and save as a dataframe
  area <- data.frame((zonal(cell_size*suitable_raster, 
                                      z = eez_raster, 
                                      fun ="sum", 
                                      na.rm = TRUE)))
  
  # Join to EEZ vector to create geo-dataframe to map
  map_df <- left_join(west_coast_eez, area, by = 'rgn') %>%
  select('rgn', 'area', 'area_km2') %>%
  mutate(area_rounded = round(area),
         Percent = round(area/area_km2*100, 2)) %>%
  mutate(labels = paste0(rgn, ": ", "\n", area_rounded, " km^2"))
  
  # Map suitable area by EEZ region
  suitability_map <- tm_shape(avg_sst) +
    tm_raster(palette = "Greys",
            title = "Mean SST (C)") +
    tm_shape(map_df) +
    tm_polygons(title = "Suitable Area \n(km^2)",
              col = 'area_rounded',
              palette = "Purples",
              legend.show = TRUE) +
  tm_text("labels", size = 1/2) +
  tm_layout(main.title = paste("Suitable", species_name, "Habitat"),
          legend.position = c("right", "top"),
          bg.color = "white",
          legend.outside = FALSE) +
  tm_compass(position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"))
  map_name <- paste0(species_name, "_map")
  assign(map_name, suitability_map)
  
  return(suitability_map)
}
```

## Use Function
```{r}
#| code-summary: Expand code
suitability("Oyster", 11, 30, 0, 70)
suitability("Abalone", 8, 18, 0, 24) 
```

## Data Sources

| Data | Citation | Link |
|------------------|------------------|------------------|------------------|
| Species Requirements for Red Abalone | Bisby, F.A., M.A. Ruggiero, K.L. Wilson, M. Cachuela-Palacio, S.W. Kimani, Y.R. Roskov, A. Soulier-Perkins and J. van Hertum. 2005. (via SeaLifeBase) | [https://www.sealifebase.ca/summary/Haliotis-rufescens.html](https://www.sealifebase.ca/summary/Haliotis-rufescens.html) |
| Sea Surface Temperature Data | NOAA Coral Reef Watch. 2018, updated daily. Daily Global 5km Satellite Sea Surface Temperature Anomaly (Version 3.1, released August 1, 2018). College Park, Maryland, USA: NOAA Coral Reef Watch | [https://coralreefwatch.noaa.gov/product/vs/data.php](https://coralreefwatch.noaa.gov/product/vs/data.php) | 
| Bathymetry Data | GEBCO Compilation Group (2024) GEBCO 2024 Grid (doi:10.5285/1c44ce99-0a0d-5f4f-e063-7086abc0ea0f) | [https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area) |
| EEZ Boundary | MarineRegions. nd. EEZ Boundaries. | [https://www.marineregions.org/eez.php](https://www.marineregions.org/eez.php) |

# References
Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. Nature Ecology & Evolution, 1, 1317-1324 (2017) [https://www.nature.com/articles/s41559-017-0257-9](https://www.nature.com/articles/s41559-017-0257-9)

# Acknowledgements
This analysis was adopted from [EDS223: Geospatial Analysis & Remote Sensing Homework #4](https://eds-223-geospatial.github.io/assignments/HW4.html). Thank you to Ruth Oliver [@ryoliver](https://github.com/ryoliver) for preparing the data hosted on Google Drive.


